An in-depth study of Explainability and Adversarial Robustness for RNNs

### 2019-12-17 ex3/paper/paper.tex
Text changes

Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans. This recently gave increased interest in explainability methods. Also in this case, common methods are designed to work with images or tabular data but not with sequences. Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. %These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
Astonishingly, feature importance methods reveal that
features manipulable for adversarial flows,
%the that are manipulable features, that are left for modification by adversaries
are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.

@@

As expected, \gls{cw}  delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
The figure shows significant differences for detecting different attack families in the first place. However, also for generating adversarial flows some attack families are more susceptible than others. However, the results match to a high degree with our expectations. For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.

@@

As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}. High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a new metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.  We thus propose the \gls{ars}, as the average of the lower half  of minimum distances of adversarial samples. Since \gls{cw} is able to find adversarial samples with minimum distance, it can be used for finding the \gls{ars}  efficiently.

When using a large $c$ in \autoref{eq:carliniWagner}, it is necessary to reduce the gradient descent learning rate  to avoid numerical issues. However, since the distance term is not affected by $c$, it is necessary to scale the iteration count inversely with the learning rate. \gls{cw} with a large $c$ therefore needs more time than with a low tradeoff. For neural networks, the \gls{ars} can be efficiently found as follows: Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If more than half of the samples are now wrongly classified, stop. Otherwise increase the tradeoff of \gls{cw} and repeat. If after a predefined number of iterations -- e.g. 100 -- still not more than half of the samples are adversarial, also break.

The result is a number of adversarial samples and some that could not be made adversarial. If at least half of the samples are misclassified, we take 50\% of the samples, with those that required smaller modifications first. Of these samples, the distance of each adversarial sample to the original sample is taken and averaged. The resulting average distance is the \gls{ars}. The larger it is the better since then an adversary needs to modify the samples more to make them adversarial. If not more then half of the samples could be made adversarial, our metric is undefined since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples. Then the ratio of samples that could be made adversarial ($<$50\%) can be useful metric to determine the exact extent of the adversarial threat. Setting the threshold for the \gls{ars} to 50\% is arbitrary, but
 % reasonable as this way the \gls{ars} approximates the median of the distances of all samples and, additionally, marks the distance where an adversarial attack is more likely to be successful than unsuccessful.
 reasonable as outliers with very large distances are ignored and if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.
%we think that it is reasonable since if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.

@@

\autoref{fig:recall_ars} shows that certain attack types, like ``DoS Slowloris'' are easy to classify for an IDS but still are also surprisingly easy for an attacker to perform adversarial modifications to make them look benign.

@@

\item
\textbf{Multiple prediction outputs.}
Often an \gls{rnn} produces an output at each time step. When  applying explainability methods, the question arises which output to consider for the method. The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation. Not only does this approach require a substantially lower amount of computational resources compared to predictions occurring at a later time step, but we also expect the immediately occurring prediction outcome to exhibit a stronger dependence of the investigated feature, compared to a prediction outcome occurring many time steps later.
However, due to the complex decision processes of deep neural networks,
it is absolutely  uncertain if the network's final prediction is correctly depicted when following this approach.
\end{itemize}

@@

A model trained with feature dropout typically yields slightly lower accuracy than a regularly trained model, even if no features are left out (flow accuracy of 0.9943 vs.~0.9965). We thus recommend training two models: One regular one and one with feature dropout to use for the feature importance.

@@

Feature dropout might constitute a building block for methods based on Shapley values \cite{shapley_value_1953} like KernelSHAP \cite{lundberg_unified_2017}. Due to space limitations, we leave this for future work.
\subsubsection{Mutual Information}
\begin{figure}
\includegraphics[width=\columnwidth]{../plots/mutinfo_example.pdf}
\caption{Two distributions yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}
Input perturbation and feature dropout mainly address feature importance.
For example, assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution for perturbing a feature with the feature distribution itself. However, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

@@

By using a classical \gls{pdp} we would be likely to loose information due to the averaging over very different types of traffic. However, for network traffic in particular, investigating each sample individually is not possible. Hence, the conditional \gls{pdp} provides the ideal compromise for our application.

%
%\begin{figure*}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot_pdp/flows_full_no_ttl_pdp_outcomes_0_3.pickle_8_DoS - DDoS-Heartbleed"}.pdf}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot_pdp/flows_full_no_ttl_pdp_outcomes_0_3.pickle_10_Normal"}.pdf}
%
%\caption{Examples of conditional PD plots for source and destination port. We only analyze these features with the \gls{pdp} since it is not straightforward to define the \gls{pdp} for features that can vary for each packet in a flow. The y-axis shows the probability for malicious traffic.}
%\todo{Max: Remove these? Not really interesting or related to the rest of the paper. Refer to other paper where we talk about this already.}
%\label{fig:pdp}
%\end{figure*}
%
%In this variant of the \gls{pdp} that we call conditional \gls{pdp}, we don't take the whole dataset for the computation for the \gls{pdp} but only a subset that belongs to a certain class. Specifically, we take each attack class and compute the \gls{pdp} only over samples of this attack class.

@@

%AH. As ALE is removed, I case this also isnt needed
%A known issue about \glspl{pdp} in general is that dependencies of the feature under investigation and the remaining variables are not considered, hence potentially querying the neural network for feature combinations which cannot possibly occur. For a certain value of the $i$th feature, the whole dataset is used for averaging the neural network's outcome.

%For this reason, Accumulated Local Effects (ALE) plots have been proposed in literature. The basic idea for ALE is that instead of using all samples, only the closest samples are used for assessing the effects of one feature. ALE plots therefore inherently require a distance measure for finding the closest samples, which, as revealed in the above discussion, in our case is not possible in a straight-forward manner.

>>>>

Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans. This recently stirred up increased interest in explainability methods. Also in this case, common methods are designed to work with images or tabular data but not with sequences. Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. %These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
Astonishingly, feature importance methods reveal that
features manipulable for adversarial flows,
%the that are manipulable features, that are left for modification by adversaries
are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.

@@

As expected, \gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
The figure shows significant differences for detecting different attack families in the first place. Also for generating adversarial flows some attack families are more susceptible than others. However, the results match to a high degree with our expectations. For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.

@@

As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}. High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a new metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
Such a metric should be easy to compute, easy to interpret and robust with respect to outliers. We thus propose the \gls{ars} as the average distance of the adversarial samples to the original samples for the 50\% of the samples which an adversary could successfully modified and which have the smallest distances. Since \gls{cw} is able to find adversarial samples with minimum distance, it can be used for finding the \gls{ars} efficiently.

When using a large $c$ in \autoref{eq:carliniWagner}, it is necessary to reduce the gradient descent learning rate to avoid overly large steps that impede convergence. However, since the distance term is not affected by $c$, it is necessary to scale the iteration count inversely with the learning rate. \gls{cw} with a large $c$ therefore needs more time than with a low tradeoff. For neural networks, the \gls{ars} can be efficiently found as follows: Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If more than half of the samples are now wrongly classified, stop. Otherwise increase the tradeoff of \gls{cw} and repeat. If after a predefined number of iterations -- e.g. 100 -- still not more than half of the samples are adversarial, stop.

The result is a number of adversarial samples and some that could not be made adversarial. If at least half of the samples are misclassified, we sort the adversarial samples by ascending distance to the original samples. Then we take samples by ascending distance until we took 50\% of all samples. Of these samples, the distance of each adversarial sample to the original sample is taken and averaged. The resulting average distance is the \gls{ars}. The larger it is the better since then an adversary needs to modify the samples more to make them adversarial. If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples. Then the ratio of samples that could be made adversarial ($<$50\%) can be a useful metric to determine the exact extent of the adversarial threat. Setting the threshold for the \gls{ars} to 50\% is arbitrary, but
 % reasonable as this way the \gls{ars} approximates the median of the distances of all samples and, additionally, marks the distance where an adversarial attack is more likely to be successful than unsuccessful.
 reasonable as outliers with very large distances are ignored and if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.
%we think that it is reasonable since if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.

@@

\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly easy for an attacker to perform adversarial modifications to make them look benign.

@@

\item
\textbf{Multiple prediction outputs.}
Often an \gls{rnn} produces an output at each time step. When  applying explainability methods, the question arises which output to consider for the method. The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation. Not only does this approach require a substantially lower amount of computational resources compared to predictions occurring at a later time step, but we also expect the immediately occurring prediction outcome to exhibit a stronger dependence of the investigated feature, compared to a prediction outcome occurring many time steps later.
However, due to the complex decision processes of deep neural networks,
this does not hold true and a feature might influence a decision many time steps later.
\end{itemize}

@@

A model trained with feature dropout typically yields slightly lower accuracy than a regularly trained model, even if no features are left out (flow accuracy of 99.43\% vs.99.65\%). We thus recommend training two models: One regular one and one with feature dropout to use for the feature importance.

@@

Feature dropout might constitute a building block for methods based on Shapley values \cite{shapley_value_1953} like KernelSHAP \cite{lundberg_unified_2017}. 
%Due to space limitations, we leave this for future work.
\subsubsection{Mutual Information}
\begin{figure}
\includegraphics[width=\columnwidth]{../plots/mutinfo_example.pdf}
\caption{Two distributions yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}
Input perturbation and feature dropout mainly address feature importance.
For example, assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution for perturbing a feature with the feature distribution itself. However, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

@@

By using a classical \gls{pdp} we would be likely to lose information due to the averaging over very different types of traffic. However, for network traffic in particular, investigating each sample individually is not possible. Hence, the conditional \gls{pdp} provides the ideal compromise for our application.

@@
